---
title: 'Exercise 2'
author: "GG"
date: "2023-07-27"
output: 
    html_document:
        toc: true
---
This is an *R Markdown* document of **assignment solution** for the capacity building training [Getting started with R](https://github.com/twesigye10/getting_started_with_R) `SOLUTION: Exercise 2`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
# read packages
library(tidyverse)
```

1. Import the `iris` data set into a variable named `df_iris`.

```{r}
write_csv(iris, file = "../outputs/df_iris.csv")
```

- Read the imported `df_iris` data & assigning it to `df_iris_data` and use the pipe operator (*|>*) to add steps **2** to **5**.

```{r}
df_iris_data <- read_csv("../inputs/df_iris.csv",               # read in the data stored in the inputs
                                    show_col_types = FALSE)|>   # folder of the project.
    select(starts_with("Petal"), "Species") |>                  # 2. Subset the dataset to keep with
                                                                # columns that start with Petal and Species
rename_with(~sub("^Petal.", "", .x), starts_with("Petal.")) |>  # 3. Rename columns that start with Petal 
                                                                # by removing Petal. from the columns. 
                                                                # Maintaining only Length and Width
    filter(Length > 1.2) |>                                     # 4. Subset dataset to keep only rows where
                                                                # Length is greater than 1.2
    mutate(Length_Category =                                    # 5. Create a new column of Length_Category
                                                                # with ranges < 3 “less_than_3”, <= 5 
                                                                # “btn_3_and_5”, > 5 “greater_than_5”
               case_when(Length < 3 ~ "less_than_3",                     
                         Length <= 5 ~ "btn_3_and_5",
                         Length > 5 ~ "greater_than_5"))
```

6. Create a *bar graph* indicating the `Species` and colour the graph based on `Length_Category`.    
- **geom_bar color by category:**

```{r}
df_iris_data |> 
ggplot(aes(x = Species)) +                  # x or y column provided
    geom_bar(aes(fill = Length_Category)) + # uses stat_count() by default
    theme_bw()
```

7. Calculate the *mean Length* for each `Species` and store the result in the `df_mean_length variable`.

```{r}
df_mean_length <- df_iris_data |> 
    group_by(Species) |> 
    summarise(`Mean Length`  = mean(Length))
```

```{r tables}
knitr::kable(df_mean_length, caption = 'Average length of Species')
```

8. Create a *bar graph* for the result from the previous step.
- **customizing the graph:**

```{r}
ggplot(df_mean_length, aes(x = `Mean Length`, y = Species)) +
    geom_col(fill = "blue") +
    theme_bw()
```

- *customizing the graph:*

```{r}
ggplot(df_mean_length, aes(x = `Mean Length`, y = Species)) +
    geom_col(fill = "blue") +
    theme_bw() +
    theme(axis.ticks = element_blank(),
          axis.text.x = element_text(face = "bold", size=12),
          axis.text.y = element_text(face = "bold", size=12),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          panel.border = element_blank(),
          axis.title = element_text(size=12)) +
    ylab(label = "")  
```

9. Calculate the *proportions* based on `Species` and `Length_Category`. Export this result into a csv file.
- calculate the proportions:

```{r}
df_proportion_length <- df_iris_data |> 
    group_by(Species, Length_Category) |> 
    summarize(count = n()) |> 
    mutate(prop = count/sum(count))
```

```{r tables-df_iris_data}
knitr::kable(df_proportion_length, caption = 'Proportions of Species by category of length')
```

- **Write output csv data set out **

```{r}
write_csv(df_proportion_length, "../outputs/proportion_length.csv")
```

Note that the `R code` **printed** along with output for the purpose of visualization!